---
title: "Getting started with ggstratify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggstratify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ggstratify)
```

## Why bother describing the data first

The decisions that shape a study are made before any model is fitted: which
variables to adjust for, where to cut a continuous one, whether a subgroup is
even large enough to talk about. Those decisions are made by a person, and a
person who has not looked at the data is deciding from assumption rather than
from evidence. That has not changed now that a model can write the analysis
code for you; if anything it matters more, because the code arrives faster
than the understanding does.

Descriptive analysis is also where most of the work is, and most of the
repetition. You draw length of stay by treatment arm, then the same figure
for men, then for women, then for each study site. The plotting code is the
same every time; only the subset changes. Doing it by hand is slow and easy
to get wrong, and a collaborator will ask for the one cut you skipped.

`ggstratify` turns that loop into a few clicks and then hands you the code,
so the description can be repeated, shared, or produced on request. It runs
entirely on your machine: no network, no model, nothing leaves the session.
That is deliberate — the environments where patient data lives are usually
the environments where you cannot paste it into anything.

## Before you start: fix the types

`ggstratify` describes columns as they are typed. It does not guess what you
meant. Convert each column to the type you intend, with factor levels in the
order you want them read, and then hand it over:

```{r, eval = FALSE}
dat <- transform(
  dat,
  sex      = factor(sex, levels = c("Male", "Female")),
  severity = factor(severity, levels = c("Mild", "Moderate", "Severe")),
  age      = as.numeric(age)
)
```

A group stored as `1, 2, 3` will be described as a number. Factor level order
becomes the order of the panels, the figures and the axis.

This is also why the app takes an object rather than a file. There is no
upload control in the application, and `ggstratify()` does not accept a file
path. Data read from disk is typed by whatever the reader guessed, and would
then be described on those guesses -- which is exactly the mistake this
section is about. Read the file yourself, look at the result, fix the types
that are wrong, and pass that object:

```{r, eval = FALSE}
cohort <- read.csv("cohort.csv")
str(cohort)          # look before describing
ggstratify(cohort)
```

## Launching the app

```{r, eval = FALSE}
ggstratify(epi_cohort)
```

That is the only function in the package.

`epi_cohort` ships with the package: 600 simulated patients with a few
continuous measurements and several categorical variables.

```{r}
str(epi_cohort)
```

Any data frame will do -- a `tibble` and a `data.table` included -- or a
matrix. Whatever you pass is copied first: your object is never modified in
place, even though the app uses `data.table` internally. The name you pass it
under is the name the generated code refers to it by, so the code on the
**R-code** tab runs as it is read.

## The layers

A description has a variable being described and, around it, the variables
you want to see it within. Those are the layers, and the **Layers** panel
asks you where to put each one.

| Layers | What you get |
|---|---|
| 1 | The variable on its own: one plain figure. |
| 2 | The second variable becomes the panels of a `facet_wrap()`. Still one figure. |
| 3 or more | One figure cannot hold them. Choose which variable stays as the panels; the rest split the figures. |

With columns `A`, `B`, `C` and `D`, you might describe `D`, panel it by `C`,
and split the figures by `A` and `B`. The sidebar keeps a running sentence of
what the current layers add up to and how many figures they make.

### Separately or crossed

Two or more variables splitting the figures can be taken one at a time, which
is the default, or crossed. Ticking `sex` (2 levels) and `site` (3) gives:

* **Each variable separately** — five figures: `sex_Male`, `sex_Female`,
  `site_Site A`, `site_Site B`, `site_Site C`.
* **Every combination** — six figures: `sex_Male__site_Site A` and so on, for
  the combinations that actually occur.

Separately is the default because it is usually the question you are asking:
"what does this look like within each level of each variable", not "what does
every combination look like". Crossing two five-level variables gives
twenty-five mostly empty figures. The Strata tab tells you how many of the
possible combinations occur, so you can see the emptiness before you commit
to it.

## Missing values are excluded, once, and counted

A row that does not say which panel or which figure it belongs to cannot be
drawn in one. As soon as a variable becomes a layer, rows missing it are
excluded — from everything, in one place, before anything is counted. That is
what keeps three numbers equal: the `N` on the Strata tab, the `N` in the
figure title, and the rows actually drawn.

They are not dropped quietly. The Strata tab says `excluded rows with a
missing sex (40)`, and the generated script prints the same count when you
run it:

```{r, eval = FALSE}
layer_vars <- c("sex", "severity")
for (v in layer_vars) {
  n_na <- sum(is.na(dt[[v]]))
  if (n_na > 0L) {
    message("Excluded ", n_na, " row(s) with a missing ", v, ".")
  }
  dt <- dt[!is.na(dt[[v]])]
}
```

Missing values in the variable being *described* are left to `ggplot2`, which
warns about them in its usual way.

## Every stratum is counted

The **Strata** tab lists every figure the current layers produce, with the
number of observations behind it, and the figure titles carry the same number.
So does every panel inside a figure: panelling by `site` gives strips reading
`site: Site A (N = 296)` rather than `Site A`, so two panels of visibly
different width are never compared without saying what they were drawn from.
The variable is named on the strip as well as the level, because a level does
not always explain itself: the panels of a categorized variable are bare
ranges, and `[18,53]` alone does not say what the range is a range of.

Two kinds of stratum are reported but not drawn:

```{r}
table(epi_cohort$site)
```

`Site D` is a declared factor level that recruited nobody. Stratifying by
`site` lists it with `N = 0` and writes no figure for it — an empty figure
would be worse than an explicit zero.

```{r}
table(epi_cohort$severity)
```

`Severe` is small. At the default minimum of 10 observations it is listed
without a file name rather than drawn. Lower **Minimum N per figure** to 0 to
include it; `N = 0` strata stay excluded regardless, because there is nothing
to draw.

## Categorizing a continuous variable

"Length of stay by age group" needs an age group, and the data has an age.
The **Categorize** panel makes one, in three ways:

* **Quantiles** — equal-sized groups. Four gives quartiles.
* **Equal-width bins** — equal-width, unequal-sized.
* **Custom cut points** — the boundaries you type. `65` gives `(-Inf, 65]`
  and `(65, Inf]`, which is usually what a clinical cut-off means.

The derived variable is an ordinary factor from that point on. It appears in
every selector, and, because it is categorical, in the **Stratify** list too:
categorize `age` at 65, tick `age_cat`, and you get one figure per age group.

The generated script builds the same column with the same line, immediately
after reading the data:

```{r, eval = FALSE}
dt <- as.data.table(epi_cohort)

# Categorized continuous variables.
dt[, age_cat := cut(age, breaks = c(-Inf, 65, Inf))]
```

That is not a re-implementation of what the app did; it is the line the app
evaluated to make the column you were looking at. Removing the rule removes
the column and the line together.

Your own object is never touched — the derived column lives only in the
copy the app holds.

## Change over time

Pick **Line** as the type of graph, put time on the X axis and the
measurement on the Y axis. Time can be a visit number, a day count or a date;
unlike a scatter plot, a line plot is not held to a continuous X.

**One line per** decides what a line is. Name the subject identifier and you
get one line per subject -- the spaghetti plot you draw when you want to see
how individual patients moved, not just where the average went. Leave it
empty when each time point already carries one value, such as a mean per
visit, and the points are joined in x order within each colour group.

Either kind of figure can carry a LOWESS smoother, and so can a scatter plot:

```{r, eval = FALSE}
p <- ggplot(d, aes(x = visit, y = crp, colour = treatment)) +
  geom_line(aes(group = id), alpha = 0.6) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.75, se = FALSE) +
  theme_bw()
```

Note where the ID is mapped. It groups `geom_line()` and nothing else, so the
smoother is fitted through the subjects rather than along each of them. The
smoother does inherit the colour grouping, which is what you want: one fit per
treatment arm, drawn only from that arm's rows, in the same way that
everything else in this package is computed within the stratum it is drawn in.

The span control is the neighbourhood `stats::loess()` fits over -- larger is
smoother. The 95% band is optional and off by default, because a descriptive
smoother is a shape, not a claim.

Previews of a line plot with an ID are sampled one whole subject at a time
rather than row by row: dropping rows from inside a trajectory would draw a
line nobody has.

## Kaplan-Meier curves

Pick **Kaplan-Meier curve** as the type of graph and choose two variables: the
follow-up time and the event indicator. `epi_cohort` carries both.

```{r}
with(epi_cohort, table(death, treatment))
```

The event indicator must be `0`/`1`, `1`/`2` or `TRUE`/`FALSE`, which is what
`survival::Surv()` accepts; a two-level factor is refused with a message
rather than a stack trace. Optional extras are a 95% confidence band,
censoring marks, a y axis clamped to 0–1, and a number-at-risk table.

The risk table is drawn under the curve, one row per group, counted at the
times the x axis is labelled at — the tail of a survival curve is read very
differently when twelve people are left than when three hundred are. Curve and
table are stacked with `patchwork::wrap_plots()` and share one x axis, so a
count always sits under the point of the curve it belongs to.

The fit is computed separately within every grouping variable, every facet
and every stratum. A curve drawn in one panel is therefore estimated only
from the rows in that panel, which is the only thing it could honestly mean.

For the same reason, survival curves are the one figure type that is never
previewed from a subsample. The other types show the shape of a
distribution, which survives sampling; a survival curve *is* the estimate,
and a sampled one is a different, wider curve.

The generated script writes out the helper it uses, so it depends on
`survival` and `data.table` but not on `ggstratify`:

```{r, eval = FALSE}
km <- km_data(d, "fu_days", "death", by = c("treatment"))

p <- ggplot(km, aes(x = .time, y = .surv, colour = treatment)) +
  geom_step(linewidth = 0.8) +
  geom_point(data = function(x) x[x$.ncens > 0, ], shape = 3, size = 2,
             show.legend = FALSE) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Time", y = "Survival probability") +
  theme_bw()
```

## Looking at part of an axis

One outlier at 400 days flattens every box in the figure. **Appearance** takes
a range for each axis — **X axis from / to** and **Y axis from / to** — and
either end of either range can be left blank for "as far as the data goes".

The range zooms the figure; it does not filter it:

```{r, eval = FALSE}
p <- ggplot(d, aes(x = treatment, y = los_days)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 19) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_bw()
```

`coord_cartesian()` is the whole point of the layer. `ylim(0, 30)` or a scale
limit would drop the rows outside the range *before* the boxplot was computed,
which moves the median and the whiskers of what you are shown; the description
would then be of 96% of the data while the tab still said `N = 500`. Zooming
leaves the summary alone and looks at part of it.

A range typed here also replaces a survival curve's **Y axis from 0 to 1**,
and a Kaplan-Meier x range is passed to the number-at-risk table too, so the
counts stay under the piece of the curve on screen.

## The generated code

The **R-code** tab shows the `ggplot2` code for the figure you are looking at,
and nothing else. There is no export loop and no `ggsave()` call: writing the
files is what the **Export all figures** button is for, and a script that also
did it would bury the few lines you actually want to lift into your own
analysis.

What is there is everything the figure genuinely needs, in the order it has to
run: the categorized columns, the exclusion of rows that cannot be placed in a
panel, the subset that makes this figure the stratum it is, and any helper the
figure defines for itself.

```{r, eval = FALSE}
# ---------------------------------------------------------------------
# Generated by ggstratify
# The ggplot2 code for the figure shown on the Plot tab.
# Figure: sex: Male (N = 291)
# ---------------------------------------------------------------------
library(data.table)
library(ggplot2)

dt <- as.data.table(epi_cohort)   # <- your data

# Rows with no value for a layer variable cannot be placed in a panel or
# a figure, so they are excluded here -- once, before anything is counted.
layer_vars <- c("sex", "severity")
for (v in layer_vars) {
  n_na <- sum(is.na(dt[[v]]))
  if (n_na > 0L) {
    message("Excluded ", n_na, " row(s) with a missing ", v, ".")
  }
  dt <- dt[!is.na(dt[[v]])]
}

# The rows this figure is drawn from: sex = Male.
d <- dt[as.character(sex) == "Male"]

add_facet_n <- function(d, var) {
  d <- data.table::copy(data.table::as.data.table(d))
  v <- as.character(d[[var]])
  lv <- if (is.factor(d[[var]])) levels(d[[var]]) else sort(unique(v))
  sizes <- as.integer(table(factor(v, levels = lv)))
  strips <- sprintf("%s (N = %d)", lv, sizes)
  d[, .facet_label := factor(strips[match(v, lv)], levels = strips)]
  d[]
}

d <- add_facet_n(d, "severity")

p <- ggplot(d, aes(x = treatment, y = los_days)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 19) +
  facet_wrap(~ .facet_label) +
  labs(title = "sex_Male (N = 291)") +
  theme_bw()

p
```

When several figures are shown at once, the header says which one the code is
for; the **Which figure** selector chooses it.

## Writing the files

**Export all figures** writes every stratum the Strata tab lists, into the
output folder, using the file names that tab shows. Two formats are offered:

* **PNG**, drawn through `ragg`, at the width, height and resolution you set.
* **SVG**, a vector format: it stays sharp at any size and can be opened in
  Illustrator or Inkscape afterwards, which is what a journal usually asks
  for. Written through `svglite` when that is installed, and otherwise through
  the cairo device that ships with R.

### Why the code always matches the figure

The preview, the **Export all figures** button and the **R-code** tab all
evaluate output from the same code generator. There is no separate "drawing"
path that could drift away from the "code" path, so the code you copy
reproduces exactly what you were looking at.

## Speed

Data handling is `data.table` throughout, which is the reason the app stays
usable on data that a `dplyr`-based GUI would struggle with. On two million
rows and eight columns: the variable scan takes about a second and happens
once, a single-stratum preview takes a third of a second, and an
eighteen-panel preview takes about three seconds.

Two mechanisms do most of that work. Only the columns a figure actually reads
are extracted before plotting, and previews of very large data are drawn from
a sample — taken *within* each stratum so that small strata keep all their
rows, and one whole subject at a time when a line plot has an ID.

The caps are deliberately high: one million rows for scatter and line plots,
where every row becomes its own mark, ten million for the types that summarise
their rows before drawing anything, and never for survival curves. Sampling is
the one place where the preview stops matching the data, so it should be
reached by data that genuinely will not draw rather than by data that is
merely large — a cohort of a million rows draws its own boxplot, and gets to.
Sampling affects the preview only in any case: exported figures and the
generated code always use every row, and the app says so on screen whenever
sampling is active.

## One function

There is one exported function, `ggstratify()`, and no modes to choose
between. Add no layers and you have the single-figure explorer —
the equivalent of `ggplotgui::ggplot_shiny()`, with the same figure types,
themes and palettes, plus the survival curve and the line plot. Add layers
and the same app describes them.

Nor is there a preview mode to choose. Whether the figures are shown all at
once as panels or one at a time follows from the variables: the switch
appears only when the figures are split and no variable has claimed the
panels, which are the conditions under which both answers exist.

It is an independent implementation, not a wrapper: `ggstratify` does not
depend on `ggplotgui` and contains none of its code. It is nevertheless
released under GPL-3, the licence of the package whose figure types it
follows.
