---
title: "Getting started with ggstratify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggstratify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ggstratify)
```

## What is ggstratify?

Descriptive analysis often means making the same figure repeatedly for
different subgroups. `ggstratify` makes this easier: choose the variables
you want to describe and stratify by, and the app creates the figures for
you. It also gives you the R code used to create each figure.

Everything runs locally. Your data are not uploaded anywhere.

## Before you start

`ggstratify` uses your variables as they are typed. Make sure categorical
variables are factors and that their levels are in the order you want.

```{r, eval = FALSE}
dat <- transform(
  dat,
  sex = factor(sex, levels = c("Male", "Female")),
  severity = factor(severity, levels = c("Mild", "Moderate", "Severe")),
  age = as.numeric(age)
)
```

The app takes a data object, not a file. Read your data first, check the
variables, and then pass the object to `ggstratify()`:

```{r, eval = FALSE}
cohort <- read.csv("cohort.csv")
str(cohort)
ggstratify(cohort)
```

Your original data are never modified.

## Launch the app

```{r, eval = FALSE}
ggstratify(epi_cohort)
```

`epi_cohort` is a simulated dataset included with the package.

```{r}
str(epi_cohort)
```

The same function works with any data frame, tibble, data.table, or matrix.

## Stratifying your figures

The **Layers** panel controls how figures are split.

* **No layer:** one figure.
* **One panel variable:** several panels in one figure.
* **Additional stratification variables:** separate figures.

For example, you could describe `D`, show panels by `C`, and create separate
figures by `A` and `B`.

When several stratification variables are selected, you can either:

* **Separate:** make figures for each variable independently.
* **Cross:** make a figure for every combination of their levels.

The **Strata** tab shows the figures that will be produced and their sample
sizes before you export them.

## Missing values and small strata

Rows missing a value for a stratification variable are excluded because they
cannot be assigned to a figure or panel. The number of excluded rows is shown
in the app and in the generated code.

Strata with fewer observations than the selected **Minimum N per figure** are
listed but not plotted. Set the minimum to `0` if you want to include small
strata. Empty factor levels are never plotted.

## Deriving a variable

The **Derive a variable** panel makes a new categorical variable out of one
you already have, using:

* **Quantiles**
* **Equal-width bins**
* **Custom cut points**
* **Missing vs observed**

For example, you can create an age group at 65 years:

```{r, eval = FALSE}
dt[, age_cat := cut(age, breaks = c(-Inf, 65, Inf))]
```

The new variable can then be used like any other categorical variable.

### Missing vs observed

The first three methods read a continuous variable's values. The fourth reads
only whether there is a value, so it applies to a variable of any type -- a
factor, a character column or a date as much as a number:

```{r, eval = FALSE}
dt[, crp_missing := factor(is.na(crp), levels = c(FALSE, TRUE),
                           labels = c("Observed", "Missing"))]
```

Stratify by it to describe another variable within each group: how the people
whose CRP was never measured differ from the people whose CRP was. Every row
lands in one of the two groups, so none is excluded -- the rows that a layer
variable's own missingness would have dropped are the ones this puts on the
screen.

Describing the variable itself within its own missingness is the one
arrangement to avoid: the Missing figure would be drawn from rows that have
no value for it. The app says so under **Layers** when you ask for that.

A variable with no missing values, or with nothing but missing values, gives
one group rather than two and is refused with a note saying which.

`crp` in the bundled `epi_cohort` is left incomplete on purpose so that this
can be tried straight away: 122 of the 600 patients were never tested, and the
milder the presentation the more often the test was skipped.

## Figure types

`ggstratify` supports several common descriptive plots, including:

* Boxplots
* Histograms
* Bar plots
* Scatter plots
* Line plots
* Kaplan-Meier curves

For line plots, choose a time variable and measurement, and optionally
specify an ID to draw one line per subject. A LOWESS smoother can also be
added.

For Kaplan-Meier curves, select the follow-up time and event variables.
Optional confidence intervals, censoring marks, and risk tables are available.

All plots are calculated within the selected panels and strata.

## Changing the axis range

**Appearance** lets you set the X- and Y-axis ranges.

The range only zooms the figure; it does not remove observations from the
analysis. This means that summary statistics such as boxplot medians remain
unchanged.

## Generated R code

The **R-code** tab shows the code used to create the figure currently shown.

The code includes the necessary data preparation, stratification, and
`ggplot2` commands, so you can copy it into your own analysis.

For example:

```{r, eval = FALSE}
library(data.table)
library(ggplot2)

dt <- as.data.table(epi_cohort)

d <- dt[sex == "Male"]

p <- ggplot(d, aes(x = treatment, y = los_days)) +
  geom_boxplot() +
  theme_bw()

p
```

The preview, exported figures, and R-code tab all use the same code generator,
so the generated code reproduces the figure shown in the app.

## Exporting figures

Click **Export all figures** to save the figures listed in the **Strata** tab.

You can export:

* **PNG** for raster images
* **SVG** for vector graphics

The exported figures use the full dataset, even when a large dataset is
sampled for the on-screen preview.

## In short

The basic workflow is:

1. Prepare and check your data.
2. Run `ggstratify(your_data)`.
3. Choose the variable to describe.
4. Add panels or stratification variables if needed.
5. Adjust the appearance.
6. Check the **Strata** tab.
7. Copy the R code or export the figures if needed.
